Mau A: A brief but intensively occupied Late Pleistocene open-air site on the Red River, northern Vietnam

### 2024-03-13 analysis/paper/paper.qmd
writing the intro

---
title: "Title Goes Here"
author:
  - Pham Thanh Son:
      correspondence: "no"
      email: thanhson9119@gmail.com
      orcid: 0000-0003-1689-0557
      institute:
        - IofA
  - Quinn Habedank:
      correspondence: "no"
      institute: UofW
  - Seungki Kwak:
      correspondence: "no"
      institute: fosg
      orcid: 0000-0003-1689-0559
  - Ben Marwick:
      correspondence: "yes"
      orcid: 0000-0001-7879-4531
      institute: UofW
institute:
  - UofW:
      name: University of Washington 
      address: Seattle, WA, USA
  - IofA: Vien Khao co hoc
title-block-published: "Last updated"  
date: now
date-format: long
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Flaked stone artefacts from Southeast Asia typically lack visually distinctive and strongly patterned forms, which can make them challenging to analyze and interpret. As a result, many of the cultural dynamics of Pleistocene hunter-gatherers of this region are poorly understood. We use 2D shape data to hypothesize a relationship between unretouched flake shape and assemblage reduction intensity at Mau A, an early Holocene archaeological site in northern Vietnam. We apply a Principal Components Analysis to the flake landmarks to investigate shape variation throughout the reduction sequence (measured by dorsal cortex coverage). We find that flake shape varies by reduction stage, primarily through differences in flake length and width. Our results suggest that flake shape is sensitive to assemblage reduction intensity, and may give useful comparative insights when other attributes show little variation. These results are important for understanding stone artefact assemblages from Southeast Asia which often yield little variation when analysed with traditional approaches.
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

@@

<!-- BM: In this section we need to review in very precise detail, previous work on investigating flake shape and identifying what affects flake shape, especially the Borel papers, not just in Southeast Asia, but anywhere in the world, Look in Borel to see what papers they cite, and we should summarise the most relevant of those, e.g. the previous papers that are most similar to our project  -->

<!-- also make sure to provide a general context of paleolithic southeast asia. See previously suggested papers and look at context papers cited by Borel et al.  -->

Stone artefact manufacture in Mainland Southeast Asia (MSEA) may have began as early as 800 ka ago (Forestier et al., 2022), but many of these claims are contested (Marwick et al. 2022) and it is not until the Late Pleistocene that an indisputably strong signal of artefact production is evidence. The oldest skeletal remains of anatomically modern humans date to roughly 50-70 ka ago (Demeter et al., 2017; Shackelford et al., 2018). By the Last Glacial Maximum at MIS 2 many areas of the region were occupied by stone artefact using forager groups. In Northern Vietnam this time was characterized by sporadic human occupation of caves, localized differences in climate (with shifting cover of rain forest between the moist highlands and the variable lowlands) and the accumulation of mollusk shell middens (McAdams et al., 2020; 2022). Throughout the Late Pleistocene humans in MSEA would repeatedly reuse caves, accumulating numerous layers of discarded shells, ash, and lithics over multiple periods of occupation, with intervening periods of abandonment and occupation by faunal species such as bats (McAdams et al., 2020).

Lithic technologies in the wider Southeast Asian region during this time are characterized by flaked stone tools which rarely have strongly patterned forms (Mijares 2008; Borel et al. 2013; Borel et al. 2017). This lack of clear typo-technological categories for flakes makes understanding the cultural dynamics of period difficult, relative to regions with more visually distinctive technologies such as northern Europe and parts of Africa and South America. In their attempt to ... Borel et al. (2013; 2017) have employed multivariate analysis of flake outlines in relation to other variables, such as the traces of use wear and retouching. They used elliptical Fourier analysis to approximate artefact shape outlines as sums of sine and cosine waves, with the coefficients from these sums being passed through Principal Component Analysis (Borel et al. 2013; 2017). This allowed the outlines of stone tools to be reduced and projected onto a two-dimensional plane, facilitating comparisons between tool morphology and other variables, such as use wear. Borel et al. (2017) choose to use of outlines rather than traditional linear measurements because the claims that often fail to cover the full variation in shape often seen in stone tools.

@@

<!-- saving this for later:  Here is a citation [@Marwick2017]  -->
## Excavations at Mau A

@@

This dense lithic layer contained the majority of stone artefacts recovered from the site. The  chaîne opératoire of Mau A lithic technology is composed of unifacial shaping on long cobbles to produce sumatraloids (sumatralith-like pieces). The chaîne opératoire also involved the shaping of thick ovoid cobbles for the production of choppers or chopping-tools and half-cobbles (longitudinally split) that are shaped into tools with transverse cutting edges. Flake scar surfaces and edges are fresh and unweathered, indicating _in situ_ production and limited post-depositional movement. While previous work has claimed the Mau A assemblage is Son Vian (a precurser to Hoabinhian technology, characterised by unifacial flaking of blocky, cubic cobbles)

@@

<!-- BM: basic stats on raw materials and change in technology by depth to satisfy readers about the basic details of the assemblage,

<!-- BM: then QH's more complex analysis on flakes after the basic data -->

<!-- BM: then previous more complex analysis on cores after QH's analysis of flakes -->

@@

source(here::here("analysis/code/999-functions.R"))

>>>>

---
title: "Mau A: A brief but intensively occupied Late Pleistocene open-air site on the Red River, northern Vietnam"
author:
  - Pham Thanh Son:
      correspondence: "no"
      orcid: 0000-0003-1689-0557
      institute:
        - IofA
  - Quinn Habedank:
      correspondence: "no"
      institute: UofW
  - Seungki Kwak:
      correspondence: "no"
      institute: fosg
      orcid: 0000-0003-1689-0559
  - Ben Marwick:
      correspondence: "yes"
      orcid: 0000-0001-7879-4531
      institute: UofW
institute:
  - UofW:
      name: Department of Anthropology, University of Washington 
      address: Seattle, WA, USA
  - IofA: 
      name: Vien Khao co hoc [Vietnamese Institute of Archaeology]
      address: Hanoi, Vietnam
title-block-published: "Last updated"  
date: now
date-format: long
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  We report on a stone artefact assemblage from Mau A, an open air site by the Red River in northern Vietnam. Artefacts from Mau A, like many sites in mainland Southeast Asia  lack visually distinctive and strongly patterned forms, which can make them challenging to analyze and interpret. As a result, many of the dynamics of Pleistocene hunter-gatherers of this region are poorly understood. We explore the use of artefact shape data to investigate relationships between  shape and assemblage reduction intensity at Mau A, a brief but intensively occupied site in northern Vietnam. We apply a Principal Components Analysis to  two dimensional landmarks of unretouched flakes to investigate shape variation throughout the reduction sequence (measured by dorsal cortex coverage). We find that flake shape varies by reduction stage, primarily through differences in flake length and width. We use three dimensional shape analysis of cores to explore their variability. Our results suggest that flake shape is sensitive to assemblage reduction intensity, and may give useful comparative insights when other attributes show little variation. These results are important for understanding stone artefact assemblages from Southeast Asia which often yield little variation when analysed with long-established approaches.
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

@@

While the timing of first stone artefact technologies in mainland Southeast Asia (mSEA) is still uncertain, with claims of over 0.7 Ma [@dereviankoDiscoveryBifacialIndustry2018] contested on geoarchaeological grounds [@marwickTektiteGeoarchaeologyMainland2021], there is general consensus on the charater of Pleistocene lithic technologies as dominated by flaked stone tools which rarely have strongly patterned forms [@mijaresMinoriCaveExpedient2002; @borelHowInterpretInformal2013]. This lack of widely-recognised typo-technological categories for artefacts has slowed progress in understanding the cultural dynamics, relative to regions with more visually distinctive technologies, such as northern Europe and parts of Africa and South America. Much recent work in mSEA has focussed on detailed qualitative documentation of small numbers of cores [e.g. @forestierReductionSequencesHoabinhian2022], with the far more numerous flake component often left unexamined. Meanwhile, novel approaches to quantitative morphometric analyses of lithic assemblages in other regions has demonstrated potential of these methods to address key questions about past human behaviour [e.g. @Okumura_Araujo_2014; @riedeQuantitativeAnalysisFinal2024; @archerGeometricMorphometricRelationship2018]. In island Southeast Asia, applications of this approach have been led by Borel et al. [-@borelHowInterpretInformal2013; -@borelStoneToolForms2017] demonstrating the potential for quantiative analysis of flake shape on assemblages from Song Terus, Indonesia.

Borel et al. [-@borelHowInterpretInformal2013; -@borelStoneToolForms2017] employed multivariate analysis of flake outlines in relation to other variables, such as traces of usewear and retouch. They used elliptical Fourier analysis to quantify artefact shape outlines, and summarised shape differences and similarities with Principal Component Analysis. This allowed the outlines of stone tools to be reduced and projected onto a two-dimensional plane, facilitating comparisons between tool morphology and other variables. Borel et al. use outlines rather than traditional linear measurements because these often fail to cover the full variation in shape often seen in stone tools. They found that there was no difference in shape between retouched and unretouched artefacts, and no correlation between a specific overall shape and any specific function indicated by usewear analysis [@borelHowInterpretInformal2013; @borelStoneToolForms2017].

We build on this pioneering work to explore relationships between shape and lithic reduction in mSEA. In some traditions of lithic analysis, reduction sequence analysis can mean reconstructing a detailed narrative of a single artefact or handful or artefacts, based on the sequence of flake removals indicated by flake scars on the artefact [e.g. @yinghuaNewTechnologicalAnalysis2021 ]. There are two important limitations to this approach. First is that it is wasteful because the majority of the assemblage (e.g. the unretouched flakes) are excluded and the information contained in these pieces is not utlized. Ethical fieldwork should prioritize maximizing information gained while minimizing destruction. Second is that it historically has shown limited potential to provide data to engage with broader anthropological questions about relationships between technology, mobility and risk in hunter-gatherer groups. Tackling these questions has been been highly productive in many parts of the world [@holdawaySurfaceStoneArtifact2020; @goldsteinHuntergathererTechnologicalOrganization2022] To address these limitations, our concept of lithic reduction follows Shott "To improve its practice, TO must treat curation as a continuous variable not categorical state, assimilate the time-averaged quality of assemblages as accumulations, reconcile the synchronic behavior it explains with the diachronic accumulations it analyzes, accommodate history as well as adaptation, and transform itself from narrative device to a body of specified, predictive theory. "

reduction analysis in mSEA is sometimes Inspired by this work, we explore shape-based methods for characterising flake and core variability in mSEA.

@@

<!-- saving this for later:  Here is a citation [@Marwick2017]  -->

## Excavations at Mau A

@@

This dense lithic layer contained the majority of stone artefacts recovered from the site. The chaîne opératoire of Mau A lithic technology is composed of unifacial shaping on long cobbles to produce sumatraloids (sumatralith-like pieces). The chaîne opératoire also involved the shaping of thick ovoid cobbles for the production of choppers or chopping-tools and half-cobbles (longitudinally split) that are shaped into tools with transverse cutting edges. Flake scar surfaces and edges are fresh and unweathered, indicating *in situ* production and limited post-depositional movement. While previous work has claimed the Mau A assemblage is Son Vian (a precurser to Hoabinhian technology, characterised by unifacial flaking of blocky, cubic cobbles)

@@

```{=html}
<!-- BM: basic stats on raw materials and change in technology by depth to satisfy readers about the basic details of the assemblage,

<!-- BM: then QH's more complex analysis on flakes after the basic data -->
```
<!-- BM: then previous more complex analysis on cores after QH's analysis of flakes -->

<!-- Eric's work: https://docs.google.com/presentation/d/1BYQBAb2gunTpRa4Z_DaTjbebYQolxIxL/edit#slide=id.p1 -->

<!-- Quinn's work: https://docs.google.com/presentation/d/1kjQ2BLKRSzCvrn6idHLBcRKfuQwczyA5/edit#slide=id.p6 https://drive.google.com/drive/u/1/folders/1tPZNYUZ0WTwWakB_-voJXRcVcTO5iWVO -->

@@

source(here::here("analysis/paper/code/999-functions.R"))
